---
title: "EDA Proposal" 
author: "DELETEMOREDATA"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
library(corrplot)
library(GGally)
library(car)

install.packages("car")

oilData <- read.csv("clean.csv")

# Converting qualitative variables to factors
oilData$region <- as.factor(oilData$region)
oilData$LL <- as.factor(oilData$LL)
oilData$qual_score <- as.factor(oilData$qual_score)

head(oilData)
str(oilData)
```

\newpage

## Introduction

There are three main research questions that we are trying to answer. Each of which is very similar, with the difference being between the predictors. The research questions are as follows:

1.  Does a higher GDP lead to a higher percentage of electricity derived from oil?

2.  Does a higher population lead to a higher percentage of electricity derived from oil?

3.  Does a higher percentage of a country's economy in industry lead to a higher percentage of electricity derived from oil?

## Data Summary

We compiled global country-level economics data from the World Bank Global Development Indices Dataset. We combined this with information from the Word Nuclear Association, Transparency International, and Britannica. In total our dataset has 55 country-level observations measuring yearly averages for the year 2023. We have recorded Percentage of Electricity Derived from Oil Sources, GDP, Population, Percentage of GDP from Industrial Sources, Net trade, Number of Nuclear Reactors, Region, Landlocked Status, and Corruption level.

In our dataset, we do not have any NA's. We dropped any country with NA values for concerns regarding data accuracy. We also dropped five extreme outliers well over a standard deviation away from the rest of our values. We are trying to evaluate the impact of various levels on developed and developing economies. All five of the outliers were below that threshold.

From the histogram created for the response variable, the data appears to be continuous and quantitative. This indicates that it is suitable for regression. For the majority of the quantitative predictors (trade, industry_pct, ln_gdp sqrt_reactors), there appears to be a negative correlation between them, while ln_population is the only one with a positive correlation. For the qualitative variables, there appears to be a difference between each of their levels. For region, regions 1 (Asia) and 2 (Europe) the lowest average electricity oil percent while regions 3 (North and South America) and 4 (Africa) have the highest. For the landlocked variable, it seems that countries that are not landlocked have a slightly larger average electricity oil percentage than countries that are. For the corruption score, there seem to be large differences. Countries with a higher level of corruption (1) have a much higher usage of their electricity with oil than countries with a lower corruption score (3).

Finally, there were three interaction plots created. The first interaction is trade and whether or not a country is landlocked. The slopes differ, suggesting there could possible be an interaction between the two. The second interaction is trade and region. Between regions 3 and 4, there does not appear to be much of an interaction, and the same applies to regions 1 and 2; however, if these pairs of regions are grouped together, their slopes differ greatly. The third interaction is ln_gdp and corruption level. For countries with a corruption level of 1, the slope is greatly different from the other two levels. For corruption groups 2 and 3, there is not too much of a difference in slope. There could be an interaction.

The VIF results indicate that multicollinearity is generally low in the model explaining electricity generated from oil (%). While ln_population and ln_gdp show moderate correlation, their VIF values remain below concerning thresholds, suggesting they do not severely distort the estimated effects. Overall, the predictors appear sufficiently independent to support reliable inference about factors influencing electricity from oil.

### Metadata Description

### EDA Summary

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this code chunk can be used to execute any functions or summaries 
## that will not be included in your output such as correlations or 5 number summaries

model <- lm(electricity_oil_pct ~ industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + trade + qual_score, data = oilData)

vif(model)

cor(oilData[, c("trade","industry_pct","ln_gdp","sqrt_reactors", "ln_population")],
    oilData$electricity_oil_pct, use="pairwise.complete.obs")

```

## Analysis and Results

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output

```

## Conclusions

\newpage

## Appendix A: Data Dictionary

| Variable Name | Abbreviated Name | Description |
|:----------------:|:-------------------:|:-------------------------------:|
| Gross Domestic Product | GDP | Market value of goods and services of a country in a specific period |
| trade | trade | This is the second variable. |
| Electricity Oil Percentage | electricity_oil_pct | percentage of electricity from oil |
| Industry Percentage | industry_pct | percentage of GDP from industry |
| Nuclear Reactors | Reactors | Number of nuclear reactors in particular country |
| Landlocked | LL | Binary variable: 1 if landlocked, 0 if not |
| Region | Region | Geographic Regions, 1: Asia, 2: Europe, 3: Americas, 4: Africa, 5: Oceania |
| Corruption level | qual_score | Degree of corruption |
| population | population | Population of a country |

\newpage

## Appendix B: Data Rows

```{r, echo=FALSE}
head(oilData)
```

\newpage

### Appendix C: Exploratory Data Analysis

```{r, echo=FALSE}
## Use this code  to include the relevant plots from EDA
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

hist(oilData$electricity_oil_pct,
     xlab = "Percentage of Electricity from Oil",
     main = "Histogram of Response Variable",
     breaks = 5,       # increase number of bins
     xaxt = "n")        # suppress x-axis

# Add custom x-axis ticks every 5%
axis(side = 1, at = seq(0, 100, by = 1))  # 0%, 5%, 10%, ..., 100%

quant_vars <- names(oilData)[sapply(oilData, is.numeric)]
quant_vars <- setdiff(quant_vars, "electricity_oil_pct")

for (var in quant_vars) {
  
  df <- oilData[, c(var, "electricity_oil_pct")]
  df <- df[complete.cases(df), ]
  
  plot(df[[var]], df$electricity_oil_pct,
       xlab = var,
       ylab = "Electricity from Oil (%)",
       main = paste("Electricity Oil % vs", var))
  
  model <- lm(electricity_oil_pct ~ x, data = transform(df, x = df[[var]]))
  abline(model, col = "red")
}

# By region
ggplot(oilData, aes(x = region, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Region")

# By landlocked
ggplot(oilData, aes(x = LL, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Landlocked Status")

# By corruption level
ggplot(oilData, aes(x = qual_score, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightcoral") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Corruption Level")

ggplot(oilData, aes(x = trade,
y = electricity_oil_pct,
color = factor(LL))) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "Trade × Landlocked Interaction",
x = "Trade",
y = "Electricity from Oil (%)",
color = "Landlocked")

ggplot(oilData, aes(x = trade,
y = electricity_oil_pct,
color = region)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "Trade × Region Interaction",
x = "Trade",
y = "Electricity from Oil (%)")

ggplot(oilData, aes(x = ln_gdp,
y = electricity_oil_pct,
color = qual_score)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "GDP × Institutional Quality Interaction",
x = "Log GDP",
y = "Electricity from Oil (%)")

# This partition is optional, but can be useful for a simple way to resize your plots
par(mfrow=c(2,3),pin=c(1,1),mar=c(7.1,4.1,3,1))
```

\newpage

## Appendix D: Final Model Output and Plots

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from assumptions and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots
```

\newpage

## Appendix E: References

### Background

We complied global country-level economics data from the World Bank Global Development Indices Dataset. We combined this with information from the Word Nuclear Association, Transparency International, and Britannica. Our total dataset records Percentage of Electricity Derived from Oil Sources, GDP, Population, Percentage of GDP from Industrial Sources, Net trade, Number of Nuclear Reactors, Region, Landlocked Status, and Corruption level data.

### Data Sources
The Editors of Encyclopaedia Britannica. (n.d.). Landlocked country. Encyclopaedia Britannica. https://www.britannica.com/topic/landlocked-country

Transparency International. (2023). Corruption perceptions index 2023. https://www.transparency.org/en/cpi/2023

United Nations Statistics Division. (n.d.). Statistical methodology and classifications. https://unstats.un.org/

World Bank. (2024a). Electricity production from oil sources (% of total) [Data set]. World Development Indicators. https://data.worldbank.org/indicator/EG.ELC.PETR.ZS

World Bank. (2024b). GDP (current US$) [Data set]. World Development Indicators. https://data.worldbank.org/indicator/NY.GDP.MKTP.CD

World Bank. (2024c). Population, total [Data set]. World Development Indicators. https://data.worldbank.org/indicator/SP.POP.TOTL

World Bank. (2024d). Trade (% of GDP) [Data set]. World Development Indicators. https://data.worldbank.org/indicator/BN.GSR.GNFS.CD

World Nuclear Association. (n.d.). World nuclear power reactors and uranium requirements. https://world-nuclear.org/information-library/facts-and-figures/world-nuclear-power-reactors-and-uranium-requireme

### Additional Help
