---
title: "EDA Proposal" 
author: "DELETEMOREDATA"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
library(corrplot)
library(GGally)
library(car)
library(MASS)
library(dplyr)

install.packages("car")

oilData <- read.csv("clean.csv")

# Converting qualitative variables to factors
oilData$region <- as.factor(oilData$region)
oilData$LL <- as.factor(oilData$LL)
oilData$qual_score <- as.factor(oilData$qual_score)

head(oilData)
str(oilData)
```

\newpage

## Introduction

There are three main research questions that we are trying to answer. Each of which is very similar, with the difference being between the predictors. The research questions are as follows:

1.  Do countries with a higher GDP have a higher percentage of electricity derived from oil?

2.  Do countries with a higher population have a higher percentage of electricity derived from oil?

3.  Do countries with a higher corruption level have a higher percentage of electricity derived from oil?

## Data Summary

In our dataset, we do not have any NA's. We dropped any country with NA values for concerns regarding data accuracy. We also dropped five extreme outliers well over a standard deviation away from the rest of our values. We are trying to evaluate the impact of various levels on developed and developing economies. All five of the outliers were below that threshold.

From the histogram created for the response variable, the data appears to be continuous and quantitative. This indicates that it is suitable for regression. For the majority of the quantitative predictors (trade, industry_pct, ln_gdp sqrt_reactors), there appears to be a negative correlation between them, while ln_population is the only one with a positive correlation. For the qualitative variables, there appears to be a difference between each of their levels. For region, regions 1 (Asia) and 2 (Europe) the lowest average electricity oil percent while regions 3 (North and South America) and 4 (Africa) have the highest. For the landlocked variable, it seems that countries that are not landlocked have a slightly larger average electricity oil percentage than countries that are. For the corruption score, there seem to be large differences. Countries with a higher level of corruption (1) have a much higher usage of their electricity with oil than countries with a lower corruption score (3).

Finally, there were three interaction plots created. The first interaction is trade and whether or not a country is landlocked. The slopes differ, suggesting there could possible be an interaction between the two. The second interaction is trade and region. Between regions 3 and 4, there does not appear to be much of an interaction, and the same applies to regions 1 and 2; however, if these pairs of regions are grouped together, their slopes differ greatly. The third interaction is ln_gdp and corruption level. For countries with a corruption level of 1, the slope is greatly different from the other two levels. For corruption groups 2 and 3, there is not too much of a difference in slope. There could be an interaction.

The VIF results indicate that multicollinearity is generally low in the model explaining electricity generated from oil (%). While ln_population and ln_gdp show moderate correlation, their VIF values remain below concerning thresholds, suggesting they do not severely distort the estimated effects. Overall, the predictors appear sufficiently independent to support reliable inference about factors influencing electricity from oil.

### Metadata Description

We compiled global country-level economics data from the World Bank Global Development Indices Dataset. We combined this with information from the Word Nuclear Association, Transparency International, and Britannica. In total our dataset has 55 country-level observations measuring yearly averages for the year 2023. We have recorded Percentage of Electricity Derived from Oil Sources, GDP, Population, Percentage of GDP from Industrial Sources, Net trade, Number of Nuclear Reactors, Region, Landlocked Status, and Corruption level.

### EDA Summary

```{r, echo=FALSE, warning=FALSE,results=TRUE}
## Use this code chunk can be used to execute any functions or summaries 
## that will not be included in your output such as correlations or 5 number summaries

model <- lm(electricity_oil_pct ~ industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + trade + qual_score, data = oilData)

print(vif(model))

print(cor(oilData[, c("trade","industry_pct","ln_gdp","sqrt_reactors", "ln_population")],
    oilData$electricity_oil_pct, use="pairwise.complete.obs"))

```

## Analysis and Results

We conducted a multiple linear regression (MLR) analysis to examine the relationship between country-level characteristics and the percentage of electricity generated from oil sources. The initial full model included the following predictors: trade, industry percentage, log population, log GDP, square root of nuclear reactors, landlocked status, region, and corruption level.

We began with a full model including all explanatory variables identified during the exploratory data analysis. These variables were selected based on their relevance to our research questions and initial evidence of association with the response variable. To assess multicollinearity, we calculated Variance Inflation Factors (VIFs) for all predictors. Most VIF values were below commonly accepted thresholds, indicating that multicollinearity was not a major concern. Although log population and log GDP showed moderate correlation, their inclusion did not significantly distort coefficient estimates, so both variables were retained in the model.

A nested F-test was conducted to evaluate a potential interaction between log GDP and corruption level. The null hypothesis states that all interaction coefficients are equal to zero, while the alternative hypothesis states that at least one interaction coefficient is not zero. The test statistic follows an F distribution with 2 numerator degrees of freedom and 40 denominator degrees of freedom. The computed value of the test statistic was F = 1.7214 with a p-value of 0.1918. At the 0.05 significance level, we fail to reject the null hypothesis, indicating that there is insufficient evidence to conclude that the interaction between ln_GDP and corruption level significantly improves the model. As a result, the interaction term is not included in the final model.

A global F-test was then conducted to assess the overall significance of the model. The null hypothesis states that all regression coefficients are equal to zero, while the alternative hypothesis states that at least one coefficient is not equal to zero. This test follows an F distribution with 12 numerator degrees of freedom and 42 denominator degrees of freedom. The observed test statistic was F = 2.105 with a p-value of 0.0377. Since the p-value is less than 0.05, we reject the null hypothesis and conclude that the model—including trade, industry percentage, ln population, ln GDP, square root of reactors, LL, region, and qual_score is statistically significant in predicting a country’s percentage of electricity derived from oil.

Next, individual t-tests were conducted for the variables of interest. For ln_population, the null hypothesis states that its coefficient is equal to zero, while the alternative states that it is not equal to zero. The test statistic follows a t distribution with 42 degrees of freedom, yielding t = 1.566 with a p-value of 0.1248. Because the p-value exceeds 0.05, we fail to reject the null hypothesis and conclude that ln_population is not a statistically significant predictor of electricity generated from oil when controlling for the other variables in the model.

For ln_gdp, the null hypothesis similarly states that its coefficient is equal to zero, with the alternative stating it is not equal to zero. Using a t distribution with 42 degrees of freedom, the test statistic was t = −0.295 with a p-value of 0.7697. Again, we fail to reject the null hypothesis at the 0.05 significance level, indicating that ln_GDP is not a statistically significant predictor in the presence of the other variables.

Finally, because qual_score is a categorical variable, its significance was assessed through a joint interpretation of its individual coefficient t-tests. The null hypothesis states that all coefficients corresponding to the levels of qual_score are equal to zero, while the alternative states that at least one is not equal to zero. Each coefficient was tested using a t distribution with 42 degrees of freedom. The results were t = −0.113 with p = 0.9106 for qual_score2 and t = −0.361 with p = 0.7201 for qual_score3. Since both p-values exceed 0.05, we fail to reject the null hypothesis and conclude that corruption level, as measured by qual_score, is not a statistically significant predictor of electricity generated from oil when considered alongside the other variables in the model.

Even though none of the individual t-tests for ln_population, ln_gdp, or qual_score are statistically significant at the 0.05 level, we will retain these variables in the model due to their theoretical relevance and role in explaining cross-country differences in energy structure. These predictors may still contribute to the overall explanatory power of the model in combination with other variables, and removing them could lead to omitted variable bias or an incomplete representation of the underlying relationships.

We then evaluated the assumptions of multiple linear regression using diagnostic plots. The residual plots suggested mild violations of normality and potential heteroscedasticity. The residual vs. fitted plots had clustering of data points, and the spread of residuals was not constant. Furthermore, the normal Q-Q plot showed deviations from linearity, especially in the upper tail, suggesting a violation of normality. Additionally, one high-leverage observation was identified, but was retained due to no evidence of data error.

To address these violations, we applied a Box-Cox transformation to the response variable. The optimal transformation parameter ($\lambda$) was selected by maximizing the log-likelihood function. After transforming the response variable, we refit the model using the same set of predictors. The transformed model showed improved adherence to regression assumptions. Residuals appeared more normally distributed, and the variance was more constant across fitted values. This indicated that the Box-Cox transformation was appropriate and improved the validity of our model.

Once the model was fit, a recalculation of the VIFs was done to compare to the original model. VIF values remained essentially the same compared to the non-transformed model, so multicollinearity remained a non-concern.

## Conclusions

Our model predicted the following equation:

$$
\begin{aligned}
\text{Electricity percentage derived from oil} =
- 0.3693
- 0.1343\,\text{industry_pct}
+ 0.5693\,\text{ln_population} \\
- 0.07246\,\text{ln_gdp} 
- 0.3462\,\text{sqrt_reactors}
- 0.1675\,\text{LL} \\
+ 0.6346\,\text{region2}
+ 2.014\,\text{region3}
+ 1.971\,\text{region4}
+ 1.244\,\text{region5} \\
- 6.718 \times 10^{-13}\,\text{trade}
- 2.142\,\text{qual_score2}
- 3.059\,\text{qual_score3}
\end{aligned}
$$

Our model predicts that a 1% increase in the population of a country will increase the percentage of electricity derived from oil by 0.57%. Similarly, it predicts that a country in Europe has 0.63% more of their electricity production generated from oil relative to a country in Asia. GDP was inversely correlated with percentage of electricity derived from oil with a 1% increase in GDP predicted to decrease the percentage of electricity derived from oil by 0.07%.

Our model is not very useful for prediction. While our model rejected the null in the global F-test, proving it has some predictive power, our individual t-tests for ln_population, ln_gdp, and qual_score, were not statistically significant at the 0.05 level.

Predicting the percentage of electricity derived from oil for the country of Algeria:

```{r, echo=FALSE, warning=FALSE,results=TRUE}
algeria <- filter(oilData, country_name == "Algeria")

algeria_predicted <- predict(model, newdata = algeria)

algeria_predicted/10

algeria$electricity_oil_pct
```

Our model predicts that the percentage of electricity derived from oil for the country of Algeria is 25.6%. The actual value of percentage of electricity derived from oil for Algeria is 27.9%.

Predicting the percentage of electricity derived from oil for the country of Croatia:

```{r, echo=FALSE, warning=FALSE,results=TRUE}
croatia <- filter(oilData, country_name == "Croatia")

croatia_predicted <- predict(model, newdata = croatia)

croatia_predicted/10

croatia$electricity_oil_pct
```

Our model predicts that the percentage of electricity derived from oil for the country of Croatia is 20.9%. The actual value of percentage of electricity derived from oil for Algeria is 21.1%.

The three main research questions we wanted to answer (see below) were not fully addressed by our model. Our individual t-tests for ln_population, ln_gdp, and qual_score were not statistically significant at the 0.05 level. This would seem to suggest that none of these factors are significant when predicting percentage of electricity derived from oil. To further address these questions, we want to instead approach the questions of GDP and population using qualitative data. GDP, which is highly correlated with population, may have been obscured, so using development level instead could help focus our results. To further explore our initial questions, we want to add more variables to find what the main predictors of percentage of electricity derived from oil truly are.

1.  Do countries with a higher GDP have a higher percentage of electricity derived from oil?

2.  Do countries with a higher population have a higher percentage of electricity derived from oil?

3.  Do countries with a higher corruption level have a higher percentage of electricity derived from oil?

\newpage

## Appendix A: Data Dictionary

| Variable Name | Abbreviated Name | Description |
|:----------------:|:-------------------:|:-------------------------------:|
| Gross Domestic Product | GDP | Market value of goods and services of a country in a specific period |
| trade | trade | This is the second variable. |
| Electricity Oil Percentage | electricity_oil_pct | percentage of electricity from oil |
| Industry Percentage | industry_pct | percentage of GDP from industry |
| Nuclear Reactors | Reactors | Number of nuclear reactors in particular country |
| Landlocked | LL | Binary variable: 1 if landlocked, 0 if not |
| Region | Region | Geographic Regions, 1: Asia, 2: Europe, 3: Americas, 4: Africa, 5: Oceania |
| Corruption level | qual_score | Degree of corruption |
| population | population | Population of a country |

\newpage

## Appendix B: Data Rows

```{r, echo=FALSE}
head(oilData)
```

\newpage

### Appendix C: Exploratory Data Analysis

```{r, echo=FALSE}
## Use this code  to include the relevant plots from EDA
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

hist(oilData$electricity_oil_pct,
     xlab = "Percentage of Electricity from Oil",
     main = "Histogram of Response Variable",
     breaks = 5,       # increase number of bins
     xaxt = "n")        # suppress x-axis

# Add custom x-axis ticks every 5%
axis(side = 1, at = seq(0, 100, by = 1))  # 0%, 5%, 10%, ..., 100%

quant_vars <- names(oilData)[sapply(oilData, is.numeric)]
quant_vars <- setdiff(quant_vars, "electricity_oil_pct")
quant_vars <- quant_vars[-3]
quant_vars <- quant_vars[-7]
quant_vars <- quant_vars[-6]
quant_vars <- quant_vars[-5]
quant_vars <- quant_vars[-2]
quant_vars <- quant_vars[-1]

for (var in quant_vars) {
  
  df <- oilData[, c(var, "electricity_oil_pct")]
  df <- df[complete.cases(df), ]
  
  plot(df[[var]], df$electricity_oil_pct,
       xlab = var,
       ylab = "Electricity from Oil (%)",
       main = paste("Electricity Oil % vs", var))
  
  model <- lm(electricity_oil_pct ~ x, data = transform(df, x = df[[var]]))
  abline(model, col = "red")
}

# By corruption level
ggplot(oilData, aes(x = qual_score, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightcoral") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Corruption Level")

ggplot(oilData, aes(x = ln_gdp,
y = electricity_oil_pct,
color = qual_score)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "GDP × Institutional Quality Interaction",
x = "Log GDP",
y = "Electricity from Oil (%)")


# This partition is optional, but can be useful for a simple way to resize your plots
par(mfrow=c(2,3),pin=c(1,1),mar=c(7.1,4.1,3,1))
```

\newpage

## Appendix D: Final Model Output and Plots

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from assumptions and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

## Interaction Testing
reducedModel <- lm(electricity_oil_pct ~ trade + industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + qual_score, data = oilData)

completeModel <- lm(electricity_oil_pct ~ trade + industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + qual_score + qual_score * ln_gdp, data = oilData)

anova(reducedModel, completeModel)

## Assumptions and Final Model
finalModel <- lm(electricity_oil_pct ~ trade + industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + qual_score, data = oilData)

plot(finalModel)

## Box-Cox Transformation
BCTransform <- function(y, lambda=0)
{
    if (lambda == 0L) 
    { 
      log(y) 
    }
    else 
    { 
      (y^lambda - 1) / lambda 
    }
}

BCTransformInverse <- function(yt, lambda=0)
{
    if (lambda == 0L)
    { 
      exp(yt) 
    }
    else
    { 
      exp(log(1 + lambda * yt)/lambda)
    }
}

shift <- abs(min(oilData$electricity_oil_pct, na.rm = TRUE)) + 0.001
oilData$electricity_oil_pct_shifted <- oilData$electricity_oil_pct + shift

model_shifted <- lm(electricity_oil_pct_shifted ~ trade + industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + qual_score, data = oilData)

bc <- boxcox(model_shifted, lambda = seq(-2, 2, by = 0.1))
bc.power <- bc$x[which.max(bc$y)]

oilData$electricity_oil_pct.bc <- BCTransform(oilData$electricity_oil_pct_shifted, bc.power)

hist(oilData$electricity_oil_pct.bc, breaks=15, xlab = "Box Cox of Electricity Percent", main = "Histogram of Box Cox Electricity Percent From Oil"); rug(oilData$electricity_oil_pct.bc)

lmFinalModelRefit <- lm(electricity_oil_pct.bc ~ trade + industry_pct + ln_population + ln_gdp + sqrt_reactors + LL + region + qual_score, data = oilData)

summary(lmFinalModelRefit)

plot(lmFinalModelRefit)

print(vif(lmFinalModelRefit))

# This partition is optional, but can be useful for a simple way to resize your plots
par(mfrow=c(2,3),pin=c(1,1),mar=c(7.1,4.1,3,1))
```

\newpage

## Appendix E: References

### Background

We complied global country-level economics data from the World Bank Global Development Indices Dataset. We combined this with information from the Word Nuclear Association, Transparency International, and Britannica. Our total dataset records Percentage of Electricity Derived from Oil Sources, GDP, Population, Percentage of GDP from Industrial Sources, Net trade, Number of Nuclear Reactors, Region, Landlocked Status, and Corruption level data.

### Data Sources

The Editors of Encyclopaedia Britannica. (n.d.). Landlocked country. Encyclopaedia Britannica. <https://www.britannica.com/topic/landlocked-country>

Transparency International. (2023). Corruption perceptions index 2023. <https://www.transparency.org/en/cpi/2023>

United Nations Statistics Division. (n.d.). Statistical methodology and classifications. <https://unstats.un.org/>

World Bank. (2024a). Electricity production from oil sources (% of total) [Data set]. World Development Indicators. <https://data.worldbank.org/indicator/EG.ELC.PETR.ZS>

World Bank. (2024b). GDP (current US\$) [Data set]. World Development Indicators. <https://data.worldbank.org/indicator/NY.GDP.MKTP.CD>

World Bank. (2024c). Population, total [Data set]. World Development Indicators. <https://data.worldbank.org/indicator/SP.POP.TOTL>

World Bank. (2024d). Trade (% of GDP) [Data set]. World Development Indicators. <https://data.worldbank.org/indicator/BN.GSR.GNFS.CD>

World Nuclear Association. (n.d.). World nuclear power reactors and uranium requirements. <https://world-nuclear.org/information-library/facts-and-figures/world-nuclear-power-reactors-and-uranium-requireme>

### Additional Help

Rossiter, D. (n.d.). Box-Cox Transformation. Box-cox transformation. <https://www.css.cornell.edu/faculty/dgr2/_static/files/R_html/Transformations.html>
