---
title: "EDA Proposal" 
author: "DELETEMOREDATA"
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE, comment=NA, warnings=FALSE,message=FALSE)
```

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this chunk for reading in packages.
## Use this chunk for reading in the data, combining the data, and any other major data cleaning that is not relevant to show.
## Remove the ## before all code.
library(corrplot)
library(GGally)

oilData <- read.csv("clean.csv")

# Converting qualitative variables to factors
oilData$region <- as.factor(oilData$region)
oilData$LL <- as.factor(oilData$LL)
oilData$qual_score <- as.factor(oilData$qual_score)

head(oilData)
str(oilData)
```

\newpage

## Introduction

There are three main research questions that we are trying to answer. Each of which is very similar, with the difference being between the predictors. The research questions are as follows:

1.  Does a higher GDP lead to a higher percentage of electricity derived from oil?

2.  Does a higher population lead to a higher percentage of electricity derived from oil?

3.  Does a higher percentage of a country's economy in industry lead to a higher percentage of electricity derived from oil?

## Data Summary


### Metadata Description

### EDA Summary

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## Use this code chunk can be used to execute any functions or summaries 
## that will not be included in your output such as correlations or 5 number summaries

```

## Analysis and Results

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output

```

## Conclusions

\newpage

## Appendix A: Data Dictionary

| Variable Name | Abbreviated Name |         Description          |
|:-------------:|:----------------:|:----------------------------:|
| Gross Domestic Product| GDP  | Market value of goods and services of a country in a specific period |
| trade         | trade  | This is the second variable. |
| Electricity Oil Percentage  | electricity_oil_pct | percentage of electricity from oil |
| Industry Percentage   |  industry_pct  | percentage of GDP from industry |
| Nuclear Reactors  | Reactors     | Number of nuclear reactors in particular country |
| Landlocked    |   LL             | Binary variable: 1 if landlocked, 0 if not |
| Region        | Region           |Geographic Regions, 1: Asia, 2: Europe, 3: Americas, 4: Africa, 5: Oceania |
| Corruption level| qual_score     | Degree of corruption | 1:Very Corrupt 2: Moderately Corrupt, 3: Fair |
| population     | population       | Population of a country |

\newpage

## Appendix B: Data Rows

```{r, echo=FALSE}
head(oilData)
```

\newpage

### Appendix C: Exploratory Data Analysis

```{r, echo=FALSE}
## Use this code  to include the relevant plots from EDA
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots

hist(oilData$electricity_oil_pct,
     xlab = "Percentage of Electricity from Oil",
     main = "Histogram of Response Variable",
     breaks = 5,       # increase number of bins
     xaxt = "n")        # suppress x-axis

# Add custom x-axis ticks every 5%
axis(side = 1, at = seq(0, 100, by = 1))  # 0%, 5%, 10%, ..., 100%

quant_vars <- names(oilData)[sapply(oilData, is.numeric)]
quant_vars <- setdiff(quant_vars, "electricity_oil_pct")

for (var in quant_vars) {
  
  df <- oilData[, c(var, "electricity_oil_pct")]
  df <- df[complete.cases(df), ]
  
  plot(df[[var]], df$electricity_oil_pct,
       xlab = var,
       ylab = "Electricity from Oil (%)",
       main = paste("Electricity Oil % vs", var))
  
  model <- lm(electricity_oil_pct ~ x, data = transform(df, x = df[[var]]))
  abline(model, col = "red")
}

# By region
ggplot(oilData, aes(x = region, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Region")

# By landlocked
ggplot(oilData, aes(x = LL, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Landlocked Status")

# By corruption level
ggplot(oilData, aes(x = qual_score, y = electricity_oil_pct)) +
  geom_boxplot(fill = "lightcoral") +
  theme_minimal() +
  labs(title = "Electricity Oil % by Corruption Level")

ggplot(oilData, aes(x = trade,
y = electricity_oil_pct,
color = factor(LL))) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "Trade × Landlocked Interaction",
x = "Trade",
y = "Electricity from Oil (%)",
color = "Landlocked")

ggplot(oilData, aes(x = trade,
y = electricity_oil_pct,
color = region)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "Trade × Region Interaction",
x = "Trade",
y = "Electricity from Oil (%)")

ggplot(oilData, aes(x = ln_gdp,
y = electricity_oil_pct,
color = qual_score)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE) +
labs(title = "GDP × Institutional Quality Interaction",
x = "Log GDP",
y = "Electricity from Oil (%)")

# This partition is optional, but can be useful for a simple way to resize your plots
par(mfrow=c(2,3),pin=c(1,1),mar=c(7.1,4.1,3,1))
```

\newpage

## Appendix D: Final Model Output and Plots

```{r, echo=FALSE, warning=FALSE,results=FALSE}
## This code chunk can be used to execute any functions or summaries that will not be included in your output
```

```{r, echo=FALSE}
## Use this section to output the summary of your final model
## Include the relevant plots from assumptions and your additional techniques
## This should NOT be a dump of all of your plots and outputs.
## Refer to the Rmarkdown guide for sizing information on your plots
```

\newpage

## Appendix E: References

### Background

### Data Sources

### Additional Help
